---
title: "Prinstats homework 1"
output: html_notebook
---
## Questions
Consider the paper of David Spiegelhalter (available on Ufora). In this paper
the author examines the age-specific risk of dying (with or from) Covid-19 in
the UK. Table 2 in the paper, gives for several age categories the numbers
of reported Covid-19 deaths, numbers of non-covid related deaths, population
sizes, among other numbers.


## 1. Question

Assign to each age interval a numerical value given by the midpoint (e.g. for 0-4, the midpoint is 
(0+4)/2 =2) and let $p_i$ denote the corresponding
population size in the UK.
Assume that the number of non-covid deaths $(Y^N )$ within an age class
with midpoint $a_i$ $(i = 1,\ldots , 11)$ can be described by a Poisson distribution with mean given by

$$
\text{E}(Y^N|\text{age}=a_i) = p_i\exp(\beta_0^N+\beta_1^Na_i)
$$

(the superscript refers to non-covid).
Also for the covid-related deaths $(Y^C)$ we assume a Poisson distribution
with

$$
\text{E}(Y^C|\text{age}=a_i) = p_i\exp(\beta_0^C+\gamma\beta_1^Na_i)
$$

*How do you interpret the $\gamma$ parameter?*

### Solution

Besides the first and the last age intervals, the difference between the mid point from one age group to the other is 10 years. 

* Lets take a unit increase  in age as 10 years.  Then 
$$
\beta_1^N = \log \text{E}(Y^N|\text{age} = a_i+1) - \log \text{E}(Y^N|\text{age} = a_i)
$$
and 
 $$
\gamma\beta_1^N = \log \text{E}(Y^C|\text{age} = a_i+1) - \log \text{E}(Y^C|\text{age} = a_i)
$$
* $\beta_1^N$ is the expected change in the number of non-covid-19 deaths per unit increase in age in the logarithmic scale
* $\gamma \beta_1^N$ is the expected change in the number of covid-19 deaths per unit increase in age in the logarithmic scale
    
* $\gamma = \gamma\beta_1^N/\beta_1^N$ is ratio of the expected change in the number of covid-19 deaths on the expected change in the number of non-covid 19 deaths for a unit increase in age in the logarithmic scale
* $\gamma>0$ implies for a unit increase in age the expected change in the number of  covid-19 deaths is greater than that of non-covid 19 deaths.  
* $\gamma<0$ implies for a unit increase in age, the expected change in the number of  covid-19 deaths is less than that of non-covid 19 deaths. 
* $\gamma = 0$ implies for a unit increase in age, the expected change in the number of  covid-19 deaths is equal to  that of non-covid 19 deaths.  Hence dying from covid 19 compared to dying from a non-covid 19 cause is independent of age.


## 2. Question

Given the models from the previous question, and assuming that all num-
bers of deaths (both covid and non-covid) are independently distributed.  Construct the log-likelihood function and make a graph of the log-likelihood as a function of the $\gamma$ parameter. You may use the following estimates for the other parameters:

$$
\beta_0^N = −12.06 ~~ \beta_0^C = −14.15 ~~ \beta_0^N = 0.09836
$$

### Solution 
We have been told that  $Y^N$ and $Y^C$ follow a Poisson distribution, recall that if a random variable $K$ follows a Poission distribution, it has the follow mass function  

$$
P_\lambda(K=k) = f(k|\lambda) = \frac{\lambda^{k}e^{-\lambda}}{k!}
$$
Let's say we have observed $n$ values from $n$ random variables $K_1,\ldots,K_n$ that are:

* all identically distributed (i.e. they all follow a Poisson distribution with parameter $\lambda_i$, well not exactly identical since their mean $\lambda$s are different) 
* independent, i.e, $P(K_1,\ldots,K_n) = \prod_i P(K_i)$

The likelihood is then defined as a function of $\mathbf \Lambda = (\lambda_1,\ldots,\lambda_n)$ given the data, i.e.,
$$
L(\mathbf \Lambda|K_1=k_1,\ldots,K_n=k_n) = \prod_i P_{\lambda_i}(K_i=k_i) = \prod_if(k_i|\lambda_i) = \prod_i\frac{\lambda_i^{k_i}e^{-\lambda_i}}{k_i!}
$$
The log likelihood which is simpler transformation to work with is given by 
$$
l(\Lambda|K_1=k_1,\ldots,K_n=k_n) = \log \prod_i P_{\lambda_i}(K_i=k_i)  = \sum_i\log\frac{\lambda_i^{k_i}e^{-\lambda_i}}{k_i!}
$$


$$
l(\Lambda|K_1=k_1,\ldots,K_n=k_n)  = \sum_i  \big(k_i \log\lambda_i - \lambda_i  - \log k_i!\big)
$$


$$
l(\Lambda|K_1=k_1,\ldots,K_n=k_n)  = \sum_i k_i \log\lambda_i - \sum_i \lambda_i  - \sum_i \log k_i!
$$

For the problem at hand we have:

* $\lambda_i^N = \text{E}(Y^N|\text{age}=a_i)$ for non-covid 19 deads
* $\lambda_i^C (\gamma) = \text{E}(Y^C|\text{age}=a_i)$ for covid 19 deads
* our parameter of interest is  $\gamma$. 
* samples size  $n_N=11$ for the covid-19 deads 
* samples size  $n_C=11$ for the non-covid-19 deads

$$
l(\gamma|Y^C_1=Y^C_1,\ldots,Y^C_n=Y^C_n,Y^N_1=y^N_1,\ldots,Y^N_n=y^N_n)  = \log \Pi_if(Y^C|\lambda_i^C(\gamma))f(y_i^N|\lambda_i^N)
$$
$$
l(\gamma|Y^C_1=Y^C_1,\ldots,Y^C_n=Y^C_n,Y^N_1=y^N_1,\ldots,Y^N_n=y^N_n)  = \sum_{i=1}^{11}\log f(Y^C|\lambda_i^C(\gamma))+\sum_{i=1}^{11}\log f(y_i^N|\lambda_i^N)
$$
$$
l(\gamma|Y^C_1=Y^C_1,\ldots,Y^C_n=Y^C_n,Y^N_1=y^N_1,\ldots,Y^N_n=y^N_n)  = \sum_{i=1}^{11}\big(Y^C\log \lambda_i^C(\gamma) + y_i^N\log \lambda_i^N\big) -\sum_{i=1}^{11}\big(\lambda_i^C(\gamma)+\lambda_i^N\big)-\sum_{i=1}^{11}\log (Y^Cy_i^N)
$$
We can simply write our log likelihood function as 

$$
l(\gamma)  = \sum_{i=1}^{11}\big(Y^C\log \lambda_i^C(\gamma) + y_i^N\log \lambda_i^N\big) -\sum_{i=1}^{11}\big(\lambda_i^C(\gamma)+\lambda_i^N\big)-\sum_{i=1}^{11}\log (Y^Cy_i^N)
$$
So we have our log likelihood function lets code it in `R`. 



```{r}
library(dplyr)
library(tidyverse)

age_mid_point = function(x,y){
  (x+y)/2
}
x = c(0,5,15,25,35,45,55,65,75,85)
y = c(4,14,24,34,44,54,64,74,84,89)

age = c(age_mid_point(x,y),90)

number_covid_19 = c(3,3,33,128,369,1283,3476,7319,16043,10160,10790)
population = c(3535430,7159102,6988755,
               7998302,7460856,8142528,
               7019590,5906928,3476922,
               918437,528959)
number_non_covid_19 = c(848,138,470,1141,2577,6577,13565,27184,47729,30703,30703)

data = data.frame(age,number_covid_19,number_non_covid_19,population)


beta_0_C_hat = -14.15
beta_1_N_hat = 0.09836
beta_0_N_hat = -12.06



beta_0_C = beta_0_C_hat
beta_0_N = beta_0_N_hat
beta_1_N = beta_1_N_hat


log_l = function(gamma_hat){

  lambda_N = data$population*exp(beta_0_N+beta_1_N*data$age)
  lambda_C = data$population*exp(beta_0_C+gamma_hat*beta_1_N*data$age)
  
  sum_1 = data$number_covid_19*log(lambda_C) +data$number_non_covid_19*log(lambda_N)
  sum_2 = lambda_C+lambda_N
  sum_3 = log(data$number_covid_19*data$number_non_covid_19)
  
  sum_all= sum(sum_1-sum_2-sum_3)
  return(-sum_all)
}
data
```

```{r}
log_ls = c()
gammas = seq(1,1.2,length.out = 1000)
for(i in gammas){
  log_ls  = c(log_l(i),log_ls)
}

log_ls = log_ls/max(abs(log_ls))

plot(x=gammas,y=-log_ls,type="l")
abline(v=gammas[-log_ls==max(-log_ls)])
```


```{r}
paste0('optimal gamma: ',round(gammas[-log_ls == max(-log_ls)],3))
```
## 3.Question 
Use the optim (or similar) function in R for finding the maxi-
mum likelihood estimate of $\gamma$.

### Solution 
Lets look for $\gamma$ with `optim`

```{r}
# initial value of gamma
par = 0

opt = optim(par=par,
      fn= log_l)

opt
```

Can you compare the results of the manual computation with that of `optim`. They seem to be different? Please check if there are any mistakes in my computations 
```{r}
plot(x=gammas,y=-log_ls,type="l")
abline(v=gammas[-log_ls==max(-log_ls)],col='blue')
abline(v=opt$par,col='red')
```


* Note that our `log_l` function returns the `negative` log likelihood instead because `optim` is a minimisation function. 

* I checked the paper for the maximum likelihood estimator of $\gamma$, they also have $(1.128)$, the value from the `optim` function. So emm.... let's check the `optimise` function ;)
```{r}
optimize(log_l,c(-2,2))
```

## 4. Question

with the information in Table 2 and with the models and their
parameter estimates of the previous questions, estimate the following two
probabilities:
$$
\text{P} (\text{Age} = 19.5 | \text{ died of covid}) \text{ and } \text{P} (\text{Age} = 79.5 | \text{ died of covid})
$$

### Solution 
Bayes theorem: for two random variables $X$ and $Y$,
$$
P(X|Y) = \frac{P(Y|X)P(X)}{P(Y)}
$$

Hence 

$$
\text{P} (\text{Age} = 19.5 | \text{ died of covid})  = \frac{\text{P} ( \text{ died of covid} | \text{Age} = 19.5 ) \text{P} (\text{Age} = 19.5)}{\text{P}( \text{ died of covid})}
$$

* further 

$$
\text{P} ( \text{ died of covid} | \text{Age} = 19.5 ) = \frac{\text{E}(Y^C|\text{Age} = 19.5)}{\text{population size at age 19.5}}
$$

Fill the values and compute this probability, do the same for $\text{P} (\text{Age} = 79.5 | \text{ died of covid})$
